
---
title: "Asthma Analysis"
author: "Sutianyi Wen"
date: "Date"
output:
  pdf_document: default
  html_document:
    highlight: pygments
    theme: spacelab
---

```{r setup, include=FALSE}
# DO NOT ALTER CODE IN THIS CHUNK
# The very first time you use this R markdown file, you should install each of the packages below.
# The same goes for other packages you might decide to use
# Remember that you only need to install each R package once in R (ever -- unless you change computers).
# All you need to do whenever you need to use the package again (after restarting the R session),
# is to use the library function to call the package.
# For example, type install.packages("knitr") in the console to install the knitr package. 
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(dplyr)
library(ggplot2)
library(MatchIt)
library(cobalt)
library(sjPlot)
library(tab)
```

* * *

```{r,echo=FALSE}
# Enter your code for loading the data here
setwd("~/Desktop/DukeFA20/IDS702/Assignment-5")
asthma = read.table("./Asthma.txt",sep = ' ',header = TRUE)
asthma$pg = as.numeric(asthma$pg - 1)
asthma$pg = as.factor(asthma$pg)
asthma$i_sex = as.factor(asthma$i_sex)
asthma$i_race = as.factor(asthma$i_race)
asthma$i_educ = as.factor(asthma$i_educ)
asthma$i_insu = as.factor(asthma$i_insu)
asthma$i_drug = as.factor(asthma$i_drug)
asthma$i_seve = as.factor(asthma$i_seve)
asthma$i_aqoc= as.numeric(asthma$i_aqoc)
# Center the three variables
asthma$com_t <- asthma$com_t - mean(asthma$com_t)
asthma$pcs_sd <- asthma$pcs_sd - mean(asthma$pcs_sd)
asthma$mcs_sd <- asthma$mcs_sd - mean(asthma$mcs_sd)
# Relevel
asthma$i_sex <- relevel(factor(asthma$i_sex), ref = 1)
asthma$i_educ <- relevel(factor(asthma$i_educ), ref = 5)
asthma$i_seve <- relevel(factor(asthma$i_seve), ref = 3)
```

### Question 1: 

Unbalanced covariates are i_sex, i_race_1, i_race_2, i_educ_4, i_educ_5, com_t and pcs_sd. I use bal.tab to print mean differences for all predictors and levels. Predictors/levels have absolute value of mean difference greater than 0.1 are considered to be unbalanced.


|          | Type     | Diff.Un |
|----------|----------|---------|
| i_age    | Contin.  | -0.0940 |
| i_sex    | Binary   | -0.1087 |
| i_race_0 | Binary   | 0.0292  |
| i_race_1 | Binary   | -0.1965 |
| i_race_2 | Binary   | 0.2101  |
| i_race_3 | Binary   | -0.0456 |
| i_race_4 | Binary   | 0.0027  |
| i_educ_5 | Binary   | 0.1706  |
| i_educ_1 | Binary   | -0.0095 |
| i_educ_2 | Binary   | -0.0075 |
| i_educ_3 | Binary   | -0.0663 |
| i_educ_4 | Binary   | -0.1650 |
| i_educ_6 | Binary   | 0.0778  |
| i_insu_1 | Binary   | -0.0383 |
| i_insu_2 | Binary   | 0.0492  |
| i_insu_3 | Binary   | 0.0058  |
| i_insu_5 | Binary   | -0.0167 |
| i_drug   | Binary   | -0.0136 |
| i_seve_3 | Binary   | -0.0158 |
| i_seve_1 | Binary   | 0.0802  |
| i_seve_2 | Binary   | -0.0069 |
| i_seve_4 | Binary   | -0.0575 |
| com_t    | Contin.  | -0.9872 |
| pcs_sd   | Contin.  | 0.7537  |
| mcs_sd   | Contin.  | -0.0531 |


```{r,echo=FALSE}
# Enter your code for question 1 here
#bal.tab(list(treat=asthma$pg,covs=asthma[,2:11],estimand="ATT"))
#love.plot(list(treat=asthma$pg,covs=asthma[,1:11],estimand="ATT"),stars = "std")
```

### Question 2(a):

According to the rang of propensity scores for two groups, there exist 8 outliers on the left tail and 40 outliers on the right tail. Therefore, we need to drop those observations to ensure the overlap.

```{r,echo=FALSE}
cov_names <- names(asthma)
p_formula <- as.formula(paste("pg ~",
                        paste(cov_names[!cov_names %in% c("i_aqoc","pg")],
                              collapse = " + ")))
pscorereg <- glm(p_formula,
                 data = asthma, family=binomial)
pscores <- predict(pscorereg, type = "response")

#sum(pscores < max(min(pscores[asthma$pg==0]),
#             min(pscores[asthma$pg==1])))
#Next the right tails
#sum(pscores > min(max(pscores[asthma$pg==0]),
#             max(pscores[asthma$pg==1])))

#If there are "outliers",
#get row index for observations that violate overlap.
index <- which((pscores < max(min(pscores[asthma$pg==0]),
                       min(pscores[asthma$pg==1])) |
       pscores > min(max(pscores[asthma$pg==0]),
                         max(pscores[asthma$pg==1]))) == TRUE)
asthma <- asthma[-index,]; pscores <- pscores[-index]

ggplot(asthma, aes(x=pscores, fill=pg)) +
  geom_density(alpha=.3) +
  xlim(0, 1)
```



### Question 2(b):

After trying to identify potential interactions between variables, there are actually no interesting interactions can improve the quality of the matched dataset so I will stick to the original formula which includes all the variables.

After dropping 48 outliers from original dataset, covariates: i_sex, i_educ_5, i_educ_4, i_educ_6, com_t, pcs_sd and mcs_sd are still have Diff.Unit greater than 1, which means they are unbalanced. And most of the mean difference from summary are negative, which means the matched data has a even worse covariates balance. However, there is nothing we can do to improve it so we will go with what we had.

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# Enter your code for question 3 here
matchesASM <- matchit(p_formula,
                    method = "nearest", distance = "logit", data = asthma)
asthma_matcheddata <- match.data(matchesASM)
#bal.tab(list(treat=asthma_matcheddata$pg,covs=asthma_matcheddata[,1:11],estimand="ATT"))
#summary(matchesASM)
```


### Question 2(c):

The causal effect Q is -0.20, which means patients' average satisfaction scores for physician group 1 is 0.20 lower than those for physician group 0 and the standard error is 0.07. The confidence interval for the causal effects are between -0.32 and 0.13, which means we are 95% confident that patients' average satisfaction scores for physician group 1 can be 0.32 lower than those for physician group 0 but also can be also 0.13 higher than those for physician group 0. And because the 95% confidence interval contains 0, we can't conclude there is a difference in patients' satisfaction scores between physical group 1 and 0.

|                       | Value |
|-----------------------|-------|
| Q                     | -0.20 |
| Standard Error        |  0.07 |
| Conf.Intv.Lower_bound | -0.32 |
| Conf.Intv.Upper_bound |  0.13 |

```{r,echo=FALSE}
# Enter your code for question 4 here
trteffct <- mean(asthma_matcheddata$i_aqoc[asthma_matcheddata$pg==1]) - 
  mean(asthma_matcheddata$i_aqoc[asthma_matcheddata$pg==0])

p_t = mean(asthma_matcheddata$i_aqoc[asthma_matcheddata$pg==1])
p_c = mean(asthma_matcheddata$i_aqoc[asthma_matcheddata$pg==0])

se_t = (p_t*(1 - p_t))/length(asthma_matcheddata$i_aqoc[asthma_matcheddata$pg==1])
se_c = (p_c*(1 - p_c))/length(asthma_matcheddata$i_aqoc[asthma_matcheddata$pg==0])

se <- sqrt(se_t + se_c)
  
lower_bound = trteffct - 1.96*se
upper_bound =  + 1.96*se
```


### Question 2(d):

The estimated causal odds ratio is 0.32. Therefore, if a patient is from physician group 1 and everything else staying the same, the odds of being satisfactory is decreased by 68%. pg is significant.

| Predictors  |  Odds Ratios |          CI         |   p   |
|-------------|:------------:|:-------------------:|:-----:|
| (Intercept) |     0.00     |     0.00 – 14.94    | 0.129 |
| i_age       |     1.03     |     0.97 – 1.10     | 0.334 |
| i_sex [1]   |     1.08     |     0.48 – 2.45     | 0.849 |
| i_race [1]  |     9.61     |    0.32 – 291.74    | 0.194 |
| i_race [2]  |     0.00     |      0.00 – Inf     | 0.994 |
| i_race [3]  | 731240142.38 |      0.00 – Inf     | 0.991 |
| i_race [4]  |     7.22     |     0.60 – 86.84    | 0.119 |
| i_educ [2]  |  12441481.44 |      0.00 – Inf     | 0.993 |
| i_educ [3]  |     0.67     |     0.04 – 11.76    | 0.784 |
| i_educ [4]  |     1.27     |     0.14 – 11.50    | 0.830 |
| i_educ [6]  |     0.51     |     0.21 – 1.25     | 0.141 |
| i_insu [2]  |     1.25     |     0.30 – 5.31     | 0.760 |
| i_insu [5]  |     2.63     |     0.26 – 26.42    | 0.411 |
| i_drug [1]  |     41.69    |    0.56 – 3129.65   | 0.090 |
| i_seve [1]  |     2.45     |     0.61 – 9.87     | 0.208 |
| i_seve [2]  |     1.12     |     0.46 – 2.75     | 0.806 |
| i_seve [4]  |     0.36     |     0.07 – 1.87     | 0.223 |
| com_t       |     2.18     |     0.53 – 9.01     | 0.283 |
| pcs_sd      |     0.90     |     0.78 – 1.04     | 0.168 |
| mcs_sd      |     0.99     |     0.93 – 1.05     | 0.705 |
| pg [1]      |     0.32     |     0.10 – 0.99     | 0.047 |
| distance    |    553.72    | 0.00 – 164221158.76 | 0.326 |

```{r,echo=FALSE}
formula = factor(i_aqoc) ~ i_age + i_sex + i_race + i_educ + i_insu + i_drug + i_seve + 
    com_t + pcs_sd + mcs_sd + pg + distance
regmodel <- glm(data=asthma_matcheddata,formula =  formula, family = binomial)
#summary(regmodel)
#tab_model(regmodel)
```


### Question 2(e):

I used one-to-n matching here and there are still unbalanced covariates: i_educ_5, i_educ_4, com_t and pcs_sd. According to summary on matched, although there are several negative values such as i_age meaning the balance gets worse from matching, most of the covariates' Diff.Unit is till less than 0.1 so we don't have to worry about them.

The causal effect Q is -0.15, which means patients' average satisfaction scores for physician group 1 is 0.15 lower than those for physician group 0 and the standard error is 0.06. The confidence interval for the causal effects are between -0.28 and 0.13, which means we are 95% confident that patients' average satisfaction scores for physician group 1 can be 0.28 lower than those for physician group 0 but also can be also 0.13 higher than those for physician group 0. And because the 95% confidence interval contains 0, we can't conclude there is a difference in patients' satisfaction scores between physical group 1 and 0.

|                       | Value |
|-----------------------|-------|
| Q                     | -0.15 |
| Standard Error        |  0.06 |
| Conf.Intv.Lower_bound | -0.28 |
| Conf.Intv.Upper_bound |  0.13 |

The estimated causal odds ratio is 0.46. Therefore, if a patient is from physician group 1 and everything else staying the same, the odds of being satisfactory is decreased by 54%. pg is significant.

| Predictors  |  Odds Ratios |         CI        |   p   |
|-------------|:------------:|:-----------------:|:-----:|
| (Intercept) |     0.00     |    0.00 – 27.59   | 0.181 |
| i_age       |     1.02     |    0.97 – 1.09    | 0.406 |
| i_sex [1]   |     1.23     |    0.60 – 2.55    | 0.573 |
| i_race [1]  |     5.99     |   0.29 – 124.45   | 0.248 |
| i_race [2]  |     0.00     |     0.00 – Inf    | 0.994 |
| i_race [3]  | 226103326.88 |     0.00 – Inf    | 0.991 |
| i_race [4]  |     4.42     |    0.48 – 40.61   | 0.189 |
| i_educ [2]  |  16684901.47 |     0.00 – Inf    | 0.993 |
| i_educ [3]  |     0.67     |    0.06 – 7.63    | 0.746 |
| i_educ [4]  |     1.00     |    0.14 – 7.09    | 0.999 |
| i_educ [6]  |     0.45     |    0.20 – 1.05    | 0.065 |
| i_insu [2]  |     0.93     |    0.25 – 3.49    | 0.917 |
| i_insu [5]  |     1.63     |    0.19 – 13.57   | 0.653 |
| i_drug [1]  |     23.07    |   0.42 – 1269.75  | 0.125 |
| i_seve [1]  |     2.24     |    0.63 – 7.92    | 0.211 |
| i_seve [2]  |     1.22     |    0.55 – 2.72    | 0.628 |
| i_seve [4]  |     0.43     |    0.10 – 1.78    | 0.244 |
| com_t       |     1.85     |    0.53 – 6.46    | 0.337 |
| pcs_sd      |     0.92     |    0.80 – 1.04    | 0.179 |
| mcs_sd      |     0.99     |    0.93 – 1.04    | 0.616 |
| pg [1]      |     0.46     |    0.22 – 0.99    | 0.048 |
| distance    |    115.09    | 0.00 – 5028684.35 | 0.384 |

The results of using a one-to-n matching are the same the one-to-one matching method. Their causal effects both have a 95% confidence interval that contains 0 and pg is significant in both models.


```{r,echo=FALSE}
matchesASM <- matchit(p_formula,
                    method = "nearest", distance = "logit", data = asthma,replace=TRUE,ratio=5)
asthma_matcheddata <- match.data(matchesASM)
#bal.tab(list(treat=asthma_matcheddata$pg,covs=asthma_matcheddata[,1:11],estimand="ATT"))
#summary(matchesASM)

trteffct <- mean(asthma_matcheddata$i_aqoc[asthma_matcheddata$pg==1]) - 
  mean(asthma_matcheddata$i_aqoc[asthma_matcheddata$pg==0])

p_t = mean(asthma_matcheddata$i_aqoc[asthma_matcheddata$pg==1])
p_c = mean(asthma_matcheddata$i_aqoc[asthma_matcheddata$pg==0])

se_t = (p_t*(1 - p_t))/length(asthma_matcheddata$i_aqoc[asthma_matcheddata$pg==1])
se_c = (p_c*(1 - p_c))/length(asthma_matcheddata$i_aqoc[asthma_matcheddata$pg==0])

se <- sqrt(se_t + se_c)

lower_bound = trteffct - 1.96*se
upper_bound =  + 1.96*se

formula = factor(i_aqoc) ~ i_age + i_sex + i_race + i_educ + i_insu + i_drug + i_seve + 
    com_t + pcs_sd + mcs_sd + pg + distance
regmodel <- glm(data=asthma_matcheddata,formula =  formula, family = binomial)
#summary(regmodel)
#tab_model(regmodel)
```

### Question 2(f):

I think __one-to-n__ matching works better in this case because it has less unbalanced covariates compared to one-to-one matching in our example. Besides matching methodology, it's more appropriate to use __regression__ to measure the causal effects compatared to computer the average causal effect Q directly because there are unbalanced covariates in the data set.


* * *



